---
title: "DATA606 Homework Presentation"
author: "Jai Jeffryes"
date: "11/18/2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Exercise 9.3 Baby weights, Part III
We considered the 
variables \texttt{smoke} and \texttt{parity}, one at a time, in modeling birth 
weights of babies in Exercises 9.1 and 9.2. A more realistic approach to modeling infant weights is to consider all possibly related variables at once. Other variables of interest include length of pregnancy in days (\texttt{gestation}), mother's age in years (\texttt{age}), mother's height in inches (\texttt{height}), and mother's pregnancy weight in pounds (\texttt{weight}). Below are three 
observations from this data set. 
\begin{center}
\begin{tabular}{r c c c c c c c}
  \hline
      & bwt & gestation & parity  & age   & height  & weight  & smoke \\ 
  \hline
1     & 120 & 284       & 0       & 27    &  62     & 100     &   0 \\ 
2     & 113 & 282       & 0       & 33    &  64     & 135     &   0 \\ 
$\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ &  $\vdots$ & $\vdots$ & $\vdots$ &   $\vdots$ \\ 
1236  & 117 & 297       & 0       & 38    &  65     & 129     &   0 \\ 
   \hline
\end{tabular}
\end{center}

## Continued
The summary table below shows the results of a regression model for predicting 
the average birth weight of babies based on all of the variables included in 
the data set.
\begin{center}
\begin{tabular}{rrrrr}
  \hline
            & Estimate  & Std. Error  & t value   & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & -80.41    & 14.35       & -5.60     & 0.0000 \\ 
gestation   & 0.44      & 0.03        & 15.26     & 0.0000 \\ 
parity      & -3.33     & 1.13        & -2.95     & 0.0033 \\ 
age         & -0.01     & 0.09        & -0.10     & 0.9170 \\ 
height      & 1.15      & 0.21        & 5.63      & 0.0000 \\ 
weight      & 0.05      & 0.03        & 1.99      & 0.0471 \\ 
smoke       & -8.40     & 0.95        & -8.81     & 0.0000 \\ 
  \hline
\end{tabular}
\end{center}

(a) Write the equation of the regression model that includes all of the variables.

## (a) Write the equation of the regression model that includes all of the variables.
\begin{tabular}{rr}
  \hline
            & Estimate \\ 
  \hline
(Intercept) & -80.41 \\ 
gestation   & 0.44 \\ 
parity      & -3.33 \\ 
age         & -0.01 \\ 
height      & 1.15 \\ 
weight      & 0.05 \\ 
smoke       & -8.40 \\ 
  \hline
\end{tabular}

$\widehat{baby\_weight}$ = -80.41 + 0.44 $\times$ $gestation$ - 3.33 $\times$ $parity$ - 0.01 $\times$ $age$ + 1.15 $\times$ $height$ + 0.05 $\times$ $weight$ - 8.40 $\times$ $smoke$

## (b) Interpret the slopes of gestation and age in this context.
$\widehat{baby\_weight}$ = -80.41 + 0.44 $\times$ $gestation$ - 3.33 $\times$ $parity$ - 0.01 $\times$ $age$ + 1.15 $\times$ $height$ + 0.05 $\times$ $weight$ - 8.40 $\times$ $smoke$

- The gestation coefficient of 0.44 says that an additional day in a pregnancy increases a baby's birth weight, on average, by 0.44 ounces.
- The age coefficient of -0.01 says that an additional year in a mother's age reduces a baby's birth weight, on average, by 0.01 ounces.

## (c) The coefficient for parity is different from the linear model shown in Exercise 9.2. Why might there be a difference?

### Single-variable model from Exercise 9.2
\begin{center}
\begin{tabular}{rrrrr}
  \hline
            & Estimate  & Std. Error    & t value   & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 120.07    & 0.60        & 199.94    & 0.0000 \\ 
parity      	& -1.93     	  & 1.19        & -1.62       & 0.1052 \\ 
  \hline
\end{tabular}
\end{center}

### Parity variable from this multiple regression model
\begin{center}
\begin{tabular}{rrrrr}
  \hline
            & Estimate  & Std. Error  & t value   & Pr($>$$|$t$|$) \\ 
  \hline
parity      & -3.33     & 1.13        & -2.95     & 0.0033 \\ 
  \hline
\end{tabular}
\end{center}

A difference like this may be due to correlation between predictor variables absent from the single-variable regression model, which are therefore confounding variables. Correlated predictor variables are said to be collinear.

## (d) Calculate the residual for the first observation in the data set.
Use the equation from answer (a). Supply the variables with values from the first observation.

$\widehat{baby\_weight}$ = -80.41 + 0.44 $\times$ $gestation$ - 3.33 $\times$ $parity$ - 0.01 $\times$ $age$ + 1.15 $\times$ $height$ + 0.05 $\times$ $weight$ - 8.40 $\times$ $smoke$

\begin{center}
\begin{tabular}{r c c c c c c c}
  \hline
      & bwt & gestation & parity  & age   & height  & weight  & smoke \\ 
  \hline
1     & 120 & 284       & 0       & 27    &  62     & 100     &   0 \\ 
   \hline
\end{tabular}
\end{center}

```{r echo=TRUE}
bwt_pred = -80.41 + 0.44*284 - 3.33*0 - 0.01*27 + 1.15*62 + 0.05*100 - 8.40*0
bwt_obs <- 120
bwt_residual <- bwt_obs - bwt_pred
```

- Observed baby weight: `r bwt_obs` ozs.
- Predicted: `r bwt_pred` ozs.
- Residual: `r bwt_residual` ozs.

## (e) The variance of the residuals is 249.28, the variance of the birth weights is 332.57. Calculate $R^2$ and adjusted $R^2$. There are 1,236 observations.

\begin{align*}
R^2 =
    1 - \frac{\text{variability in residuals}}
        {\text{variability in the outcome}}
\end{align*}

\begin{align*}
R_{adj}^{2} =
    1 - \frac{\text{variability in residuals}}
        {\text{variability in the outcome}}
		\times \frac{n-1}{\text{degrees of freedom}}
\end{align*}

Degrees of freedom: $n$ - $predictor\_count$ - 1
```{r echo=TRUE}
r2 <- 1 - (249.28 / 332.57)
n <- 1236; df <- n - 6 - 1
r2_adj <-  1 - ((249.28 / 332.57) * ((n - 1) / df))
```
- $R^2$: `r r2`
- $R^2_{adj}$: `r r2_adj`