---
title: "Chapter 2, Page 52"
author: "Jai Jeffryes"
date: "2/26/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Monte Carlo
I was curious to try the Monte Carlo simulations in [Grinstead] because of my interest in Bayesian Analysis. Outside of CUNY, I participate in a weekly study group that is learning Bayesian Analysis from the book, *Statistical Rethinking*, by Richard McElreath and the videos of his complete course on YouTube. We started simulations. Later, we'll get to Markov Chain Monte Carlo.

The [Grinstead] book has accompanying programs, and I read the Monte Carlo program in its TrueBasic version. It's easy. I wrote a version in R that utilizes vectors instead of a loop.

```{r}
MonteCarlo <- function(trials) {
	x <- runif(trials, min = 0, max = 1)
	y <- runif(trials, min = 0, max = 1)
	
	is_under_curve <- sum(y <= x^2)
	area <- is_under_curve / trials
	
	return(area)
}

options(digits = 3)
trials <- 10000

# Expected error.
error <- (1 / sqrt(trials)) * c(-1, 1)
print(error)

# From calculus, the area is 1 / 3.
area <- 1 / 3
print(area)

# Estimate
set.seed(42) # The answer to Life, the Universe, and Everything.
estimate <- MonteCarlo(trials)
print(estimate)

# Confidence interval
print(estimate + error)
```

This estimate falls within the confidence interval.

Let's take on some of the related exercises.

## Ex. 2.3
Alter the program MonteCarlo to estimate the area of the circle of radius 1/2 with center at (1/2,1/2) inside the unit square by choosing 1000 points at random. Compare your results with the true value of $\pi/4$. Use your results to estimate the value of $\pi$. How accurate is your estimate?

```{r}
MonteCarlo_2_3 <- function(trials) {
	x <- runif(trials, min = 0, max = 1)
	y <- runif(trials, min = 0, max = 1)


	is_in_circle <- sum((x^2 + y^2) <= 1)
	area <- is_in_circle / trials
	
	return(area)
}

options(digits = 3)

# From calculus, the area is pi / 4.
area <- pi / 4
print(area)

# Estimate area.
set.seed(42) # The answer to Life, the Universe, and Everything.
trials <- 1000
estimate <- MonteCarlo_2_3(trials)
print(estimate)

# True value of pi.
pi

# Estimate of pi.
pi_estimate <- estimate * 4
print(pi_estimate)

# Accuracy of pi estimate in percentage.
(1 - ((pi_estimate - pi) / pi)) * 100
```

## Ex. 2.4
Alter the program MonteCarlo to estimate the area under the graph of $y = sin \pi x$ inside the unit square by choosing 10,000 points at random. Now calculate the true value of this area and use your results to estimate the value of $\pi$. How accurate is your estimate?

```{r}
MonteCarlo_2_4 <- function(trials) {
	x <- runif(trials, min = 0, max = 1)
	y <- runif(trials, min = 0, max = 1)


	is_under_curve <- sum(y <= sin(pi * x))
	area <- is_under_curve / trials
	
	return(area)
}
```

### Area, from calculus
We need the definite integral of the curve to compute the area.

$\int_{0}^{1} sin \pi x dx$

Which equals:

$(- \frac{1}{\pi}cos \pi (1)) - (- \frac{1}{\pi}cos \pi (0))$

Or:
$\frac{2}{\pi}$

```{r}
area <- (1 - cos(pi)) / pi
print(area)

# Estimate area.
options(digits = 3)
set.seed(42) # The answer to Life, the Universe, and Everything.
trials <- 10000
estimate <- MonteCarlo_2_4(trials)
print(estimate)

# True value of pi.
pi

# Estimate of pi.
pi_estimate <- 2 / estimate
print(pi_estimate)

# Accuracy of pi estimate in percentage.
(1 - ((pi_estimate - pi) / pi)) * 100
```

## Ex. 2.5
Alter the program MonteCarlo to estimate the area under the graph of $y = 1/(x + 1)$ in the unit square in the same way as in Exercise 4. Calculate the true value of this area and use your simulation results to estimate the value of $log 2$. How accurate is your estimate?

```{r}
MonteCarlo_2_5 <- function(trials) {
	x <- runif(trials, min = 0, max = 1)
	y <- runif(trials, min = 0, max = 1)


	is_under_curve <- sum(y <= 1/(x + 1))
	area <- is_under_curve / trials
	
	return(area)
}
```

### Area, from calculus
We need the definite integral of the curve to compute the area.

$\int_{0}^{1} \frac{1}{x + 1} dx$

Which equals:

$ln(1 + 1) - ln(0 + 1)$

Which equals:

$ln(2) - 0$

Therefore, our estimate of the area under the curve is an estimate of $ln2$

```{r}
area <- log(2)
print(area)

# Estimate area.
options(digits = 3)
set.seed(42) # The answer to Life, the Universe, and Everything.
trials <- 10000
estimate <- MonteCarlo_2_5(trials)
print(estimate)

# True value of ln(2).
log(2)

# Estimate of ln(2).
ln2_estimate <- estimate
print(ln2_estimate)

# Accuracy of ln(2) estimate in percentage.
(1 - (ln2_estimate - log(2))) * 100
```
