---
title: "Discussion 10"
author: "Jai Jeffryes"
date: "3/31/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting this one
I'm delighted I can see what is going on with the material this week. I did a bunch of the exercises, and I'll probably be able to add to them even.

## p. 413, ex. 11.1
It is raining in the Land of Oz. Determine the probabilities for the next three days of weather. $w^{(1)}$, $w^{(2)}$, $w^{(3)}$

I understand this! Here is the transition matrix.
```{r}
P <- matrix(c(
	.5, .25, .25,
	.5, 0, .5,
	.25, .25, .5),
	nrow = 3, byrow = T
)
colnames(P) <- c("R", "N", "S")
rownames(P) <- c("R", "N", "S")
```

We're given a probability vector for the starting distribution. The problem says it's raining in the Land of Oz, so that entry is equal to 1.

```{r}
u = c(1, 0, 0)
```

Theorem 11.2 gives the probability that the chain is in $s_i$ after $n$ steps.

$u^{(n)} = uP^n$

```{r}
w1 <- u %*% P
w1
w2 <- u %*% P %*% P
w2
w3 <- u %*% P %*% P %*% P
w3
```

## Ex. 11.2
In example 11.4, let $a = 0$ and $b = 1/2$. Find $P$, $P^2$, and $P^3$. What would $P^n$ be? What happens to $P^n$ as $n$ tends to infinity.

I'd like a little more convenience with raising dot products multiple times.
```{r}
dot_power <- function(x, n){
	if (n < 2) stop("Invalid n")
	
	ans <- x
	for(i in 2:n){
		ans <- ans %*% x
	}
	return(ans)
}
```

```{r}
P <- matrix(c(
	1.0, 0.0,
	0.5, 0.5),
	nrow = 2, byrow = T
)
colnames(P) <- c("yes", "no")
rownames(P) <- c("yes", "no")
```

Here are $P$, $P^2$, and $P^3$.
```{r}
P
dot_power(P, 2)
dot_power(P, 3)
```

Bump $n$ up to 100 and this is the result.
```{r}
dot_power(P, 100)
```

The Markov Chain has reached a stationary state, for which the starting state is irrelevant. With enough transitions, all no answers will have been changed to yes answers, no matter what the president said..

## Ex. 11.3
In example 11.5 find $P$, $P^2$, and $P^3$. What would $P^n$ be?

```{r}
P <- matrix(c(
	.5, .25, .25,
	.5, .25, .25,
	.5, .25, .25),
	nrow = 3, byrow = T
)
colnames(P) <- c("W", "P", "S")
rownames(P) <- c("W", "P", "S")

P
dot_power(P, 2)
dot_power(P, 3)
dot_power(P, 100)
```

Answer: For all $n$, $P^n = P$

## Ex. 11.4
For Example 11.6, find the probability that the grandson of a Harvard alum went to Harvard.

```{r}
P <- matrix(c(
	.8, .2, 0,
	.3, .4, .3,
	.2, .1, .7),
	nrow = 3, byrow = T
)

statespace <- c("H", "Y", "D")
dimnames(P) <- list(statespace, statespace)

dot_power(P, 2)
```

Answer: 0.70

## Ex. 11.5
For Example 11.7, find the probability that the grandson of a Harvard alum went to Harvard.

```{r}
P <- matrix(c(
	1, 0, 0,
	.3, .4, .3,
	.2, .1, .7),
	nrow = 3, byrow = T
)

statespace <- c("H", "Y", "D")
dimnames(P) <- list(statespace, statespace)

dot_power(P, 2)
```

Answer: 1.0

## Ex. 11.6
For Example 11.9, assume we start with a hybrid bred to a hybrid. Find $u^{(1)}$, $u^{(2)}$, and $u^{(3)}$. What would $u^{(n)}$ be?

```{r}
P <- matrix(c(
	.5, .5, 0,
	.25, .5, .25,
	.0, .5, .5),
	nrow = 3, byrow = T
)

statespace <- c("GG", "Gg", "gg")
dimnames(P) <- list(statespace, statespace)

u = c(0, 1, 0)

u %*% P
u %*% dot_power(P, 2)
u %*% dot_power(P, 3)
```

Answer: for all $n$, $u{(n)} = (.25, .5, .25)$

## Ex. 11.7
Find the matrices $P^2$, $P^3$, $P^4$, and $P^n$ for the given transition matrices.

```{r}
# FIRST MATRIX
P <- matrix(c(
	1, 0,
	0, 1),
	nrow = 2, byrow = T
)

dot_power(P, 2)
dot_power(P, 3)
dot_power(P, 4)
dot_power(P, 100)

# SECOND MATRIX
P <- matrix(c(
	0, 1,
	1, 0),
	nrow = 2, byrow = T
)

dot_power(P, 2)
dot_power(P, 3)
dot_power(P, 4)
dot_power(P, 100)
dot_power(P, 101)
```

Answer:

- For the first matrix, which is the Identity matrix, $P^n = P$
- For the second, when $n$ is odd, $P^n = P$. When $n$ is even, $P^n$ = the Identity matrix.

## Supplementary article
An intro to Markov Chains in R: [https://dataconomy.com/2018/03/an-introduction-to-markov-chains-using-r/](https://dataconomy.com/2018/03/an-introduction-to-markov-chains-using-r/)

Movements of Uber drivers between zones.
```{r}
# State space.
DriverZone <- c("North", "South", "West")
# Transition matrix.
ZoneTransition <- matrix(c(0.3, 0.3, 0.4, 0.4, 0.4, 0.2, 0.5, 0.3, 0.2), nrow = 3, byrow = T)
dimnames(ZoneTransition) <- list(DriverZone, DriverZone)
ZoneTransition
```

```{r}
# install.packages("markovchain")
# install.packages("diagram")

library(markovchain)
library(diagram)

# Use the markovchain function to create a markovchain object
McZone <- new("markovchain", states = DriverZone, byrow = T,
			  transitionMatrix = ZoneTransition, name = "Driver Movement")
# Calling the object gives the transition matrix.
McZone

# Plot the transition matrix.
plotmat(ZoneTransition, pos = c(1, 2), lwd = 1, cex.txt = 0.5, box.size = 0.1,
		box.type = "circle", box.prop = 0.5, box.col = "light yellow", arr.length = .1,
		arr.width = .1, self.cex = .4, self.shifty = -.01, self.shiftx = .13,
		main = "Transition Diagram")
```

The plot can be used to find state transitions to multiply and add up. If a driver starts in the North Zone, what is the probability he'll be there after two trips?

$NN + SN + WN = 0.3*0.3 + 0.4*0.3 + 0.5*0.4$
```{r}
0.3*0.3 + 0.4*0.3 + 0.5*0.4
```

Using the Markov Chain model, we can run this code and read the NN element.

```{r}
McZone^2

# Here's three trips
McZone^3
```

Run enough trips and you eventually reach equilibrium at a stationary state, for which the starting point was irrelevant. We're there by 25 trips

```{r}
steadyStates(McZone)
McZone^25
```

If 50 drivers started in the North and completed 25 trips in a day, here's where we expect them to end up.

```{r}
50 * steadyStates(McZone)
```

