---
title: "DATA605: Homework 2"
author: "Jai Jeffryes"
date: "2/6/2020"
output: pdf_document
---

## Problem set 1
(1) Show that $A^TA \ne A A^T$ in general. (Proof and demonstration.)

### Proof
Consider two cases.

1. Let $A$ be an $m \times n$ matrix, $B$ be an $n \times p$ matrix and $m \ne n$.
  - (a) The product, $AB$, is an $m \times p$ matrix. By the definition of matrix multiplication, [Beezer, p. 179].
  - (b) Let $B = A^T$. Its dimensions are $n \times m$.
  - (c) The product, $AA^T$ is an $m \times m$ matrix. By (a) and (b).
  - (d) The product, $A^TA$ is an $n \times n$ matrix. By (a) and (b).
  - (e) Since we assumed $m \ne n$, and by (c) and (d), therefore $A^TA \ne A A^T$.
2. Assumptions.
  - Let $A$ be a square matrix, so that its dimensions $m \times n$ are given by $m = n$.
  - Define indexes for its elements such that $0 < i < m$ and $0 < j < n$.
  - Let the rows not equal the columns, such that $A_{ij} \ne A_{ji}$.
  - Therefore... *this is where I have difficulty, expressing as a proof that the calculation of the elements yields unequal products when the rows do not equal the columns, though I demonstrate it below.*

$AA^T \ne A^TA$ in these two cases.

On the other hand, in the following case $AA^T = A^TA$.

1. Let $A$ be a square matrix, so that its dimensions $m \times n$ are given by $m = n$.
  - Define indexes for its elements such that $0 < i < m$ and $0 < j < n$.
  - Let the rows equal the columns, such that $A_{ij} = A_{ji}$.
  - In this case, $AA^T = A^TA$.
  
### Demonstration
Here is a square matrix that is not symmetric.

$A = \begin{bmatrix}1 & 2\\3 & 4\end{bmatrix}$,
$A^T = \begin{bmatrix}1 & 3\\2 & 4\end{bmatrix}$

Here are the calculations worked out.

$AA^T = \begin{bmatrix}1 \cdot 1 + 2 \cdot 2 & 1 \cdot 3 + 2 \cdot 4 \\ 3 \cdot 1 + 4 \cdot 2 & 3 \cdot 3 + 4 \cdot 4 \end{bmatrix} = \begin{bmatrix}5 & 11 \\ 11 & 25\end{bmatrix}$

Computing the product in `R` confirms the correctness of the typeset demo.

```{r}
# Double check
A <- matrix(1:4, nrow = 2, byrow = TRUE)
A
A %*% t(A)
```

Reversing the terms of the product yields a different result.

$A^TA = \begin{bmatrix}1 \cdot 1 + 3 \cdot 3 & 1 \cdot 2 + 3 \cdot 4 \\ 2 \cdot 1 + 4 \cdot 3 & 2 \cdot 2 + 4 \cdot 4 \end{bmatrix} = \begin{bmatrix}10 & 14 \\ 14 & 20\end{bmatrix}$

```{r}
# Double check
t(A) %*% A
```

We can calculate in `R` the products, $AA^T$ and $A^TA$, using a matrix that is not square. This illustrates the conclusion from part #1 of the proof above, namely that the products have different dimensions.

```{r}
A <- matrix(1:6, nrow = 3, byrow = T)
A
A %*% t(A)
t(A) %*% A
```

(2) For a special type of square matrix $A$, we get $A^TA = AA^T$. Under what conditions could this be true?

**ANSWER:** This is true if a matrix is symmetric, in which case $A = A^T$. I choose a non-trivial example so I'm confident it isn't an "accident." I test the symmetry of my demo.

```{r}
A <- matrix(c(
	1,  4,  7,  0,  9,
	4,  1,  3,  6, -3,
	7,  3,  1, -2,  5,
	0,  6, -2,  1, -1,
	9, -3,  5, -1,  1
), nrow = 5, byrow = TRUE)

# Demo matrix is symmetric. All elements equal the corresponding
# transposition.
# Add up the number of unequal elements; expected result is 0.
sum(A != t(A))

# Commute their dot products and test inequality of elements, which
# should sum to 0.
t(A) %*% A
A %*% t(A)
sum(t(A) %*% A != A %*% t(A))
```

## Problem set 2
Write an R function to factorize a square matrix A into LU.

- Employ the shortcut method for LU decomposition illustrated in this week's [video](https://www.youtube.com/watch?v=UlWcofkUDDU).
- The `print()` statements got me to the iteration path I wanted.
- I wanted the design to generalize so I don't have to keep track of indexes.
- `i` and `j` retain their familiar references to row and column respectively.
- `i` and `j` walk the matrix diagonally, so they always address the location for defining the multiplier for row reduction.

```{r}
factor_lu <- function(A) {
	# Accept a matrix, A, and factor it into lower and upper
	# triangular matrices.
	# Assumptions:
	#   A is square.
	#   A row reduces to U in row echelon form without row permutation.
	
	L <- diag(nrow(A))
	for (j in 1:(ncol(A) - 1)) {
		# print(paste("j:", j))
		i <- j
		# print(paste("i:", i))
		for (k in i:(nrow(A) - 1)) {
			# print(paste("k:", k))
			multiplier <- -(A[k + 1, j] / A[i, j])
			A[k + 1, ] <- multiplier * A[i, ] + A[k + 1, ]
			L[k + 1, j] <- -multiplier
		}
	}
	
	return(list("L" = L, "U" = A))
}

# Examples from YouTube video.
matrix_1 <- matrix(c(
	1, 4, -3,
	-2, 8, 5,
	3, 4, 7
), nrow = 3, byrow = T)

matrix_2 <- matrix(c(
	2, 4, -4,
	1, -4, 3,
	-6, -9, 5
), nrow = 3, byrow = T)

matrix_1_lu <- factor_lu(matrix_1)
matrix_2_lu <- factor_lu(matrix_2)

print("SAMPLE MATRIX 1")
print(matrix_1)
print("SAMPLE MATRIX 1: LOWER AND UPPER MATRIX FACTORS")
print(matrix_1_lu)
print("COMPARE PRODUCT OF LU TO ORIGINAL MATRIX")
matrix_1_lu[[1]] %*% matrix_1_lu[[2]]

print("SAMPLE MATRIX 2")
print(matrix_2)
print("SAMPLE MATRIX 2: LOWER AND UPPER MATRIX FACTORS")
print(matrix_2_lu)
print("COMPARE PRODUCT OF LU TO ORIGINAL MATRIX")
matrix_2_lu[[1]] %*% matrix_2_lu[[2]]
```
