---
title: "DATA605 - Assignment 8"
author: "Jai Jeffryes"
date: "4/13/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4)
```

## p. 303, ex. 11
A company buys 100 lightbulbs, each of which has an exponential lifetime of 1000 hours. What is the expected time for the first of these bulbs to burn out?

We know:

1. $n = 100$. (Problem definition)
2. The expected value for each bulb's lifetime is $\mu = 1000$. (Problem definition)
3. The distribution of lifetime is exponential. (Problem definition)
4. $M$ is the minimum of the distribution of the life of a single bulb, $x_j$. (Notation)
5. $E(M) = \frac{\mu}{n}$ (Given in problem 10)
6. $E(M) = \frac{1000}{100} = 10$ (1, 2, 5)

Answer: **The expective time for the first bulb to burn out is 10 hours.**

Note: This question was all about the answers hiding in plain sight. It will be worth returning to the section to learn more about exponential distributions.

## p. 303, ex. 11
Assume that $X1$ and $X2$ are independent random variables, each having an exponential density with parameter $\lambda$. Show that $Z = X_1 − X_2$ has density

$f_Z(z) = (1/2) \lambda e^{−\lambda|z|}$

## Extra study
- [L12.2 The Sum of Independent Discrete Random Variables](https://www.youtube.com/watch?v=zbu8KQx9bqM). This is the clearest video I found explaining convolution.
- [Convolution Theorem: Application & Examples](https://study.com/academy/lesson/convolution-theorem-application-examples.html). And this is the clearest article.

### Grinstead exercises
#### Ex. 7.1, p. 289
A die is rolled three times. Find the probability that the sum of the outcomes is

(a) greater than 9.
(b) an odd number.

I'll use a solution that counts the probabilities that I found on the internet and adapted in `R`. This helped me understand the summation when I added traces.

```{r}
# PMF
P <- function(x) {
	if (x %in% c(1:6)) {
		return(1 / 6)
	} else {
		return(0)
	}
}

# S(2). Include some traces.
S2 <- function(x, verbose = F) {
	total <- 0
	
	if (verbose) {
		print(paste0("x=", x))
	}
	for (k in 0:x) {
		total <- total + (P(k) * P(x - k))

		if (verbose) {
			print(paste0("k=", k, ", x-k=", x-k))
			print(paste0("    P(k)=", P(k), ", P(x-k)=", P(x-k)))
			print(paste0("    P(k)P(x-k)=", P(k) * P(x - k)))
			print(paste0("    total=", total))
		}
	}
	return(total)
}
    
S3 <- function(x, verbose = F) {
	total <- 0
	for (k in 0:x) {
		prob <- S2(k, verbose) * P(x - k)
		total <- total + prob
	}
	return(total)
}
```

Check out the PMF
```{r}
P(0)
P(3)
P(7)
```

Check out $P(S_2 = 2)$ and compare with p. 286.

```{r}
S2(2, verbose = T)
S2(3, verbose = T)
```
Check out $P(S_2 = 2)$ and compare with p. 287.

```{r}
S3(3, verbose = T)
```

##### Answer
- Part a
```{r}
total <- 0
for (i in 1:9) {
	total <- total + S3(i)
}

1 - total
```

- Part b
```{r}
total <- 0
for (i in 0:((3 * 6) - 1)) {
	if (i %% 2 == 1) {
		total <- total + S3(i)
	}
}

1 - total
```

#### Ex. 7.2, p. 289
The price of a stock on a given trading day changes according to the distribution

$p_X= (\frac{-1}{1/4} \frac{0}{1/2} \frac{1}{1/8} \frac{2}{1/8})$

Find the distribution for the change in stock price after two (independent) trading days.

The PMF represents the change after one day. Two days of changes is the convolution of the PMF with itself.

$p_X(x = 1) \cdot p_X(x = 1$

```{r}
# PMF
P <- function(x) {
	# if (x %in% c(1:6)) {
	# 	return(1 / 6)
	# } else {
	# 	return(0)
	# }
	dist <- c(01/4, 1/2, 1/8, 1/8)
	if (x %in% c(-1:2)) {
		return(dist[x + 1])
	} else {
		return(0)
	}
}

```

