---
title: "Data 605 - Assignment 13"
author: "Jai Jeffryes"
date: "5/3/2020"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#options(scipen = 999)
```

## Introduction
There is a table of contents at the front of the document. For grading, if you do not need to see my reasoning you can jump right to the answers.

## Problem 1
Use integration by substitution to solve the integral below.

$$\int 4e^{-7x}dx$$

Let $u = -7x$, $u'(x) = -7$

$$\frac{du}{dx} = -7 \iff \frac{-1}{7} du = dx$$
$$\int 4e^{-7x}dx = 4 \int e^u \frac{-1}{7} du = \frac{-4}{7} \int e^u du$$

$$= \frac{-4}{7}(e^u + C)$$

### Answer
$$\frac{-4}{7}e^{-7x} + C$$

## Problem 2
Biologists are treating a pond contaminated with bacteria. The level of contamination is changing at a rate of $\frac{dN}{dt} = -\frac{3150}{t^4} - 220$ bacteria per cubic centimeter per day, where $t$ is the number of days since treatment began. Find a function $N(t)$ to estimate the level of contamination if the level after 1 day was 6530 bacteria per cubic centimeter.

### Approach
We're given a rate of change. We're given a value from the domain of contamination levels. We're asked for a function that returns the level of contamination.

- To do that, we'll integrate the provided derivative, producing the function for contamination.
- Express the contamination at time $t = 1$ using the provided level.
- Solve for the integration constant $C$.
- Provide the general function for contamination level.

We're given:

$$\frac{dN}{dt} = -\frac{3150}{t^4} - 220$$

So,
$$N(t) = \int (\frac{-3150}{t^4} - 220)dt$$
$$= -3150 \int t^{-4} - 220 \int dt$$
$$= -3150(\frac{t^{-3}}{-3}) - 220t + C$$
$$= \frac{1050}{t^3} - 220t + C$$

We're given:
$$N(1) = 6530$$

Substituting $t = 1$ into our antiderivative:
$$1050 - 220 + C = 6530$$
$$\iff C = 6530 - 830 = 5700$$

### Answer
Therefore:
$$N(t) = \frac{1050}{t^3} - 220t + 5700$$

## Problem 3
Find the total area of the red rectangles in the figure below, where the equation of the line is $f(x) = 2x - 9$.

```{r}
x <- c(5:8)
y <- 2 * x - 9
width <- 1

ans <- sum(width * y)
```

### Answer
`r ans`

## Problem 4
Find the area of the region bounded by the graphs of the given equations.

$y=x^2 - 2x - 2$, $y=x+2$

### Approach
- Plot the curves. Shade the area to be computed.
- Find the end points.
- Compute the definite integral under each curve and calculate their difference.

```{r}
x <-  seq(from = -5, to = 7, by = 0.1)
y.1 <-  x^2 - 2 * x - 2
y.2 <-  x + 2
shade <- x >= -1 & x <= 4

plot(x, y.1, type = "l", col = "black",
	 main = "Compute area between the curves",
	 ylab = "y")
lines(x, y.2, col = "red")
polygon(c(x[shade], rev(x[shade])), c(y.2[shade], rev(y.1[shade])),col="skyblue")
legend(4, 4, legend=c("x^2 - 2x - 2", "x + 2"),
       col=c("black", "red") , lty=c(1, 1), cex=0.8)
```

**Find the end points for the area to compute**

It's easy to discern in the plot, but let's make this honest. The end points are where the two functions are equal.

$$x^2 - 2x - 2 = x + 2$$
$$\iff x^2 - 3x - 4 = 0$$
$$= (x + 1)(x - 4)$$
$$\iff x = \{-1, 4\}$$

Guessing these roots isn't hard, but it's really easy when you've drawn the picture first.

**Integrate**
$$\int_{-1}^4 (x + 2) dx - \int_{-1}^4 (x^2 - 2x -2) dx$$
$$= \int_{-1}^4 [(x + 2) - (x^2 - 2x - 2)] dx$$
$$= \int_{-1}^4 (-(x^2) + 3x + 4) dx$$
$$= (\frac{-1}{3})x^3 + (\frac{3}{2})x^2 + 4x \bigg\rvert_{x = -1}^4$$
$$= (\frac{-1}{3})4^3 + (\frac{3}{2})4^2 + 4\cdot 4 -\frac{1}{3} - \frac{3}{2} + 4$$
```{r}
ans <- (-1 / 3) * 4^3 + (3 / 2) * 4^2 + 4 * 4 - (1 / 3) - (3 / 2) + 4
```

### Answer
`r round(ans, 2)`

I guess you can integrate in `R`, too. Who knew? Let me try this to double check my work

```{r}
integrand <- function(x) {-(x^2) + 3 * x + 4}
integrate(integrand, lower = -1, upper = 4)
```
That is pretty cool. Like magic.

Reference:

- [Shading between curves in `R`](http://www.alisonsinclair.ca/2011/03/shading-between-curves-in-r/)

## Problem 5
A beauty supply store expects to sell 110 flat irons during the next year. It costs \$3.75 to store one flat iron for one year. There is a fixed cost of \$8.25 for each order. Find the lot size and the number of orders per year that will minimize inventory costs.

### Analysis
First, I needed to figure out what "lot size" meant. I thought it had something to do with storage and I tried to figure out if enough was given about inventory costs. No, so I looked it up.

> [Definition of 'Lot Size'](https://economictimes.indiatimes.com/definition/lot-size): Lot size refers to the quantity of an item ordered for delivery on a specific date or manufactured in a single production run... A simple example of lot size is: when we buy a pack of six chocolates, it refers to buying a single lot of chocolate.

Now I remember the term "odd lots." Sellers don't want to break up a lot and sell piecemeal.

### Calculation
 - Let $x$ equal lot size. Expected sales are 110 irons, so the number of orders will be $110 / x$.
 - $Inventory\ Cost = Order\ cost + Storage\ cost = f(x)$
 - $Order\ cost = 8.25 \times \frac{110}{x}$
 - We assume our inventory is $x$ upon receipt of order, it runs down at a constant rate, and we deplete our stock before receiving another order. Therefore, we estimate the average inventory balance as the quantity ordered divided by 2. $Storage\ cost = 3.75 \times \frac{x}{2}$
 - $Inventory\ cost = 8.25 \times \frac{110}{x} + 3.75 \times \frac{x}{2}$
 
Simplified,
$$Inventory\ cost = f(x) = 907.50(x^{-1}) + 1.875(x)$$

To optimize the function, we'll find the value of x where the cost function's first derivative = 0, indicating a local extremum, and the second derivative is positive, indicating the curve is concave up and the value of x under consideration is a local minimum.

$$f'(x) = \frac{-907.50}{x^2} + 1.875$$
$$f''(x) = \frac{-2 \cdot -907.50}{x^{3}} = \frac{1815.00}{x^3}$$

Set the first derivative to 0.
$$0 = \frac{-907.50}{x^2} + 1.875$$
$$\iff -1.875 = \frac{-907.50}{x^2}$$
$$\iff x^2 = \frac{-907.50}{-1.875}$$
$$\iff x = \sqrt{\frac{907.50}{1.875}} = \pm22$$

I don't really need to consider the negative value, since you can't order negative lot sizes, but for fun, consider the second derivative for evaluating concavity.
$$f''(x) = \frac{1815.00}{x^3}$$

This calculation is negative for a negative $x$ and positive for a positive $x$. As expected, $x = 22$ is a local minimum.

I'd like to plot this and see that it behaves as expected.
```{r}
x <-  seq(from = 10, to = 50, by = 0.1)
y <- 907.50 * x^(-1) + 1.875 * x

plot(x, y, type = "l", col = "black",
	 main = "907.50 * x^(-1) + 1.875 * x")
abline(v = 22, col = "red")
```

**Reference:**

This is an accounting calculation for inventory management called [Economic Order Quantity (EOQ)](https://xplaind.com/333007/economic-order-quantity)

### Answer
5 orders of 22 flat irons per order minimize inventory costs.

## Problem 6
Use integration by parts to solve the integral below.
$$\int ln(9x) \cdot x^6 dx$$

**Variable names**

We're going to use two integration methods, both of which traditionally assign the variable name $u$ to a function. We'll call them $u_1$ and $u_2$.

### Set up parts
Integration by parts is given by:
$$\int u\ dv = uv - \int v\ du$$

Let $u_1 = ln(9x)$, $dv = x^6 dx$

### Derive remaining terms
Find the anti-derivative of $dv$
$$v = \int x^6 dx = \frac{x^7}{7} + C$$

Take the derivative of $u_1$. We apply the chain rule, designating the inner function $u_2$.
$$\frac{d}{dx}(ln(9x))$$
Let $u_2 = 9x$
$$\frac{d}{dx}(ln(9x)) = \frac{d}{du_2}(ln(u_2)) \frac{du_2}{dx} = \frac{1}{u_2} \cdot\frac{du_2}{dx}$$
Substitute for $u_2$.
$$\frac{1}{9x} \cdot 9 = \frac{1}{x}$$

Therefore,
$$du_2 = \frac{1}{x} dx$$

### Substitute into the formula for integration by parts and complete
$$\int u\ dv = uf - \int v\ du = \frac{1}{7} ln(9x) x^7 - \int \frac{x^7}{7} \cdot \frac{1}{x} dx$$
$$= \frac{1}{7} x^7 ln(9x) - \int \frac{x^6}{7} dx = \frac{1}{7} x^7 ln(9x) - \frac{x^7}{49} + C$$

### Answer
$$\frac{1}{7} x^7 ln(9x) - \frac{x^7}{49} + C$$

## Problem 7
Determine whether $f(x)$ is a probability density function on the interval [1, $e^6$]. If not, determine the value of the definite integral.
$$f(x)= \frac{1}{6x}$$

### Analysis
A PDF must have an area under the curve for the specified interval equal to 1. Let's plot this first.

```{r}
x <-  seq(from = 0, to = exp(6) + 1, by = 0.1)
y.1 <- 1 / (6 * x)
y.2 <- 0 * x
shade <- x >= 1 & x <= exp(6)

par(mfrow = c(2,2))
plot(x, y.1, type = "l", col = "black",
	 main = "Area under 1 / 6x",
	 ylab = "y")
polygon(c(x[shade], rev(x[shade])), c(y.2[shade], rev(y.1[shade])),col="skyblue")

x <-  seq(from = 0, to = 4, by = 0.1)
y.1 <- 1 / (6 * x)
y.2 <- 0 * x
shade <- x >= 1 & x <= exp(6)

plot(x, y.1, type = "l", col = "black",
	 main = "Area under 1 / 6x - Zoom left",
	 ylab = "y")
polygon(c(x[shade], rev(x[shade])), c(y.2[shade], rev(y.1[shade])),col="skyblue")

x <-  seq(from = exp(6) - 4, to = exp(6) + 1, by = 0.1)
y.1 <- 1 / (6 * x)
y.2 <- 0 * x
shade <- x >= 1 & x <= exp(6)

plot(x, y.1, type = "l", col = "black",
	 main = "Area under 1 / 6x - Zoom right",
	 ylab = "y")
polygon(c(x[shade], rev(x[shade])), c(y.2[shade], rev(y.1[shade])),col="skyblue")
```

### Integrate
$$\int_1^{e^6}\frac{1}{6x} dx$$
$$= \frac{1}{6} \int_1^{e^6} \frac{1}{x} dx$$
$$= \frac{1}{6} ln|x|  \bigg\rvert_{x = 1}^{e^6}$$
$$= \frac{1}{6} ln|e^6| - \frac{1}{6} ln|1|$$
$$= \frac{1}{6} \cdot 6| - \frac{1}{6} \cdot 0 = 1$$

Simple, but I like double checking.

```{r}
(1 / 6) * log(abs(exp(6))) - (1 / 6) * log(abs(1))
```

### Answer
Since the area under the function is equal to 1 on the interval [1, $e^6$], $f(x)$ is a probability density function.