---
title: "DATA605 - Assignment 5"
author: "Jai Jeffryes"
date: "2/27/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem set
Choose independently two numbers B and C at random from the interval [0, 1] with uniform density. Prove that B and C are proper probability distributions.
Note that the point (B,C) is then chosen at random in the unit square.
Find the probability that

### (a) B + C < 1/2.

```{r}
MonteCarlo.a <- function(trials) {
	B <- runif(trials, min = 0, max = 1)
	C <- runif(trials, min = 0, max = 1)
	
	is_under_curve <- sum(B + C < 1.0 / 2.0)
	area <- is_under_curve / trials
	
	return(area)
}

options(digits = 3)
trials <- 10000

# Expected error.
error <- (1 / sqrt(trials))
print(error)

# Estimate
set.seed(42)
estimate <- MonteCarlo.a(trials)
print(estimate)
```

**ANSWER, with 95% confidence: `r estimate` $\pm$ `r error`**

### (b) BC < 1/2.

```{r}
MonteCarlo.b <- function(trials) {
	B <- runif(trials, min = 0, max = 1)
	C <- runif(trials, min = 0, max = 1)
	
	is_under_curve <- sum(B * C < 1.0 / 2.0)
	area <- is_under_curve / trials
	
	return(area)
}

options(digits = 3)
trials <- 10000

# Expected error.
error <- (1 / sqrt(trials))
print(error)

# Estimate
set.seed(42)
estimate <- MonteCarlo.b(trials)
print(estimate)
```

**ANSWER, with 95% confidence: `r estimate` $\pm$ `r error`**

### (c) |B - C| < 1/2.

```{r}
MonteCarlo.c <- function(trials) {
	B <- runif(trials, min = 0, max = 1)
	C <- runif(trials, min = 0, max = 1)
	
	is_under_curve <- sum(abs(B - C) < 1.0 / 2.0)
	area <- is_under_curve / trials
	
	return(area)
}

options(digits = 3)
trials <- 10000

# Expected error.
error <- (1 / sqrt(trials))
print(error)

# Estimate
set.seed(42)
estimate <- MonteCarlo.c(trials)
print(estimate)

```

**ANSWER, with 95% confidence: `r estimate` $\pm$ `r error`**

### (d) max{B,C} < 1/2.

```{r}
MonteCarlo.d <- function(trials) {
	B <- runif(trials, min = 0, max = 1)
	C <- runif(trials, min = 0, max = 1)
	
	# Use pairwise maximum function.
	is_under_curve <- sum(pmax(B, C) < 1.0 / 2.0)
	area <- is_under_curve / trials
	
	return(area)
}

options(digits = 3)
trials <- 10000

# Expected error.
error <- (1 / sqrt(trials))
print(error)

# Estimate
set.seed(42)
estimate <- MonteCarlo.d(trials)
print(estimate)
```

**ANSWER, with 95% confidence: `r estimate` $\pm$ `r error`**

### (e) min{B,C} < 1/2.
 
```{r}
MonteCarlo.e <- function(trials) {
	B <- runif(trials, min = 0, max = 1)
	C <- runif(trials, min = 0, max = 1)
	
	# Use pairwise minimum function.
	is_under_curve <- sum(pmin(B, C) < 1.0 / 2.0)
	area <- is_under_curve / trials
	
	return(area)
}

options(digits = 3)
trials <- 10000

# Expected error.
error <- (1 / sqrt(trials))
print(error)

# Estimate
set.seed(42)
estimate <- MonteCarlo.e(trials)
print(estimate)
```

**ANSWER, with 95% confidence: `r estimate` $\pm$ `r error`**