---
title: "Chapter 7"
author: "Jai Jeffryes"
date: "4/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

## Extra study
- [L12.2 The Sum of Independent Discrete Random Variables](https://www.youtube.com/watch?v=zbu8KQx9bqM). This is the clearest video I found explaining convolution.
- [Convolution Theorem: Application & Examples](https://study.com/academy/lesson/convolution-theorem-application-examples.html). And this is the clearest article.

# Ex. 1, p. 289
#### Ex. 7.1, p. 289
A die is rolled three times. Find the probability that the sum of the outcomes is

(a) greater than 9.
(b) an odd number.

```{r}
P <- function(x) {
	if (x %in% c(1:6)) {
		return(1 / 6)
	} else {
		return(0)
	}
}

S2 <- function(x, verbose = F) {
	total <- 0
	for (k in 0:x) {
		total <- total + (P(k) * P(x - k))

		if (verbose) {
			print(paste0("x=", x))
			print(paste0("k=", k, ", P(k)=", P(k)))
			print(paste0("x-k=", x-k, ", P(x-k)=", P(x-k)))
			print(paste0("P(k)P(x-k)=", P(k) * P(x - k)))
			print(paste0("total=", total))
		}
	}
	return(total)
}
    
S3 <- function(x) {
	total <- 0
	for (k in 0:x) {
		prob <- S2(k) * P(x - k)
		total <- total + prob
	}
	return(total)
}

# part a
total <- 0
for (i in 1:9) {
	total <- total + S3(i)
}

1 - total

# part b
total <- 0
for (i in 0:((3 * 6) - 1)) {
	if (i %% 2 == 1) {
		total <- total + S3(i)
	}
}

1 - total
```

#### Ex. 7.1, p. 289
A die is rolled three times. Find the probability that the sum of the outcomes is

(a) greater than 9.
(b) an odd number.

I'll use a solution that counts the probabilities that I found on the internet and adapted in `R`. This helped me understand the summation when I added traces.

```{r}
# PMF
P <- function(x) {
	if (x %in% c(1:6)) {
		return(1 / 6)
	} else {
		return(0)
	}
}

# S(2). Include some traces.
S2 <- function(x, verbose = F) {
	total <- 0
	
	if (verbose) {
		print(paste0("x=", x))
	}
	for (k in 0:x) {
		total <- total + (P(k) * P(x - k))

		if (verbose) {
			print(paste0("k=", k, ", x-k=", x-k))
			print(paste0("    P(k)=", P(k), ", P(x-k)=", P(x-k)))
			print(paste0("    P(k)P(x-k)=", P(k) * P(x - k)))
			print(paste0("    total=", total))
		}
	}
	return(total)
}
    
S3 <- function(x, verbose = F) {
	total <- 0
	for (k in 0:x) {
		prob <- S2(k, verbose) * P(x - k)
		total <- total + prob
	}
	return(total)
}
```

Check out the PMF
```{r}
P(0)
P(3)
P(7)
```

Check out $P(S_2 = 2)$ and compare with p. 286.

```{r}
S2(2, verbose = T)
S2(3, verbose = T)
```
Check out $P(S_2 = 2)$ and compare with p. 287.

```{r}
S3(3, verbose = T)
```

##### Answer
- Part a
```{r}
total <- 0
for (i in 1:9) {
	total <- total + S3(i)
}

1 - total
```

- Part b
```{r}
total <- 0
for (i in 0:((3 * 6) - 1)) {
	if (i %% 2 == 1) {
		total <- total + S3(i)
	}
}

1 - total
```


[StackOverflow](https://stackoverflow.com/questions/23569133/adding-two-random-variables-via-convolution-in-r)
```{r}
f.X <- function(x) dnorm(x,1,0.5)        # normal (mu=1.5, sigma=0.5)
f.Y <- function(y) dlnorm(y,1.5, 0.75)   # log-normal (mu=1.5, sigma=0.75)
# convolution integral
f.Z <- function(z) integrate(function(x,z) f.Y(z-x)*f.X(x),-Inf,Inf,z)$value
f.Z <- Vectorize(f.Z)                    # need to vectorize the resulting fn.

set.seed(1)                              # for reproducible example
X <- rnorm(1000,1,0.5)
Y <- rlnorm(1000,1.5,0.75)
Z <- X + Y
# compare the methods
hist(Z,freq=F,breaks=50, xlim=c(0,30))
z <- seq(0,50,0.01)
lines(z,f.Z(z),lty=2,col="red")
```

#### Ex. 7.2, p. 289
The price of a stock on a given trading day changes according to the distribution

$p_X= (\frac{-1}{1/4} \frac{0}{1/2} \frac{1}{1/8} \frac{2}{1/8})$

Find the distribution for the change in stock price after two (independent) trading days.

The PMF represents the change after one day. Two days of changes is the convolution of the PMF with itself.

$p_X(x = 1) \cdot p_X(x = 1$

```{r}
# PMF
P <- function(x) {
	# if (x %in% c(1:6)) {
	# 	return(1 / 6)
	# } else {
	# 	return(0)
	# }
	dist <- c(01/4, 1/2, 1/8, 1/8)
	if (x %in% c(-1:2)) {
		return(dist[x + 1])
	} else {
		return(0)
	}
}

```

