---
title: "Multiple linear regression"
author: "Jai Jeffryes"
date: "11/27/2019"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=TRUE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.
    
**ANSWER**

This is an observational study, and there are problems with the design.

*One*

The fundamental problem with this design is the disconnection between class evaluation and beauty score, for the beauty score is assigned by someone different from the person evaluating the class.

*Two*

Two thirds of the courses have multiple instructors, yet their class evaluations are commingled as well as their beauty scores.

*Three*

There could be selection bias. The percentage of students who evaluated each class ranges from 10% to 100%.

*Four*

I would like the students evaluating the class also to record a beauty rating. However, if I accepted that, there are still for too few, beauty raters, namely six. By contrast, there are 17,000 class evaluations. Presumably, there are duplicate students among those 17,000 class evaluations, but six raters for beauty is far too few.

Rather than determining a direct connection between beauty and course evaluation, this study may observe a correlation between student course evaluations and third party perception of beauty.

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?

**ANSWER**

```{r}
hist(evals$score)
```

The distribution is symmetric, unimodal, and skewed to the left. It's central tendency is 4.5 out of a positive rating score of 5. I expect this central tendency, since I expect students to rate their instructors well. Otherwise, I expect the administration to take corrective action to achieve better performance.

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualization (scatterplot, side-by-side boxplots, or 
    mosaic plot).
    
**ANSWER**

`bty_avg` and `gender`. Women appear to inspire a higher perception of beauty than men.

```{r}
boxplot(evals$bty_avg ~ evals$gender)
```

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg, eval = TRUE}
plot(evals$score ~ evals$bty_avg)
```


Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

**ANSWER**

It seems like a leading question suggesting I'm supposed to notice something awry. However, I don't see anything. Yes, there appear to be fewer than 463 points on the plot. I looked for NA values in the data frame and there weren't any. I counted points on the plot by eye and found about 100 in the right half. All that says to me, given absence of NA values, is many points on this plot must overlap.

```{r}
sum(is.na(evals$score))
sum(is.na(evals$bty_avg))
```

4.  Replot the scatterplot, but this time use the function `jitter()` on the
    $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was
    misleading about the initial scatterplot?
    
**ANSWER**

My guess was right. Many points overlapped.
    
```{r}
plot(evals$score ~ jitter(evals$bty_avg))
```


5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the line to your plot
    using `abline(m_bty)`. Write out the equation for the linear model and
    interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
**ANSWER**

```{r}
library(openintro)
m_bty <- lm(score ~ bty_avg, data = evals)
# plot bty_avg vs. score ------------------------------------------
par(mar = c(3.5, 4, 1, 0.5), las = 1, mgp = c(2.5, 0.7, 0),
   cex.lab = 1.5, cex.axis = 1.5)
plot(score ~ bty_avg, data = evals, 
     pch = 19, col = COL[1,2], 
     xlab = "Beauty (Avg)", ylab = "Score", axes = FALSE)
axis(1)
axis(2)
# axis(2, at = seq(20, 80, 20))
box()
abline(m_bty, col = COL[2], lwd = 2)

summary(m_bty)
```

$\widehat{score}$ = 3.88 + 0.07 * $bty\_avg$

`bty_avg` appears to be a statistically significant predictor, although I noted problems with the experimental design.

`bty_avg` is not a practical predictor. The correlation is pretty flat. At a slope of 0.07, movement of a score by just a quarter point would require a difference in average beauty of almost 4. All other things being equal, an instructor with a beauty rating of 8 would be predicted to have only a quarter point higher course evaluation score than an instructor with a beauty rating of 4.

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).

**Linearity**
```{r}
plot(m_bty$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)
```

I don't see patterns in the residual plot, so it appears to satisfy the requirement of linearity.

**Nearly normal**

```{r}
hist(m_bty$residuals)
```

Residuals are skewed to the left. I don't yet understand fully when I should reject based on skewedness.

```{r}
qqnorm(m_bty$residuals)
qqline(m_bty$residuals)
```

I'm going to go with rejection. I don't think the residuals satisfy the requirement of being nearly normal.

**Constant variability**

The first plot does not present constant variability (does it?). I'm saying no, because there are is greater variance in residuals below the correlation than above. Is that the right distinction?

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel, eval = TRUE}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels, eval = TRUE}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender, eval = TRUE}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.

**Check for outliers**
```{r}
hist(m_bty_gen$residuals)
```

Residuals are normal enough and no outliers.

**Linearity and variability**

```{r}
plot(m_bty_gen$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)
```

It looks a lot like it did without the gender coefficient. Maybe this is OK.

8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?
    
**ANSWER**

Yes, `bty_avg` is still a significant predictor. Adding `gender` to the model changed the parameter estimate for `bty_avg` from 0.06664 to 0.07416.

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines, eval = TRUE}
multiLines(m_bty_gen)
```

9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?

**ANSWER**

\[
  \begin{aligned}
\widehat{score} &= 3.74734 + 0.07416 \times bty\_avg + 0.17239 \\
\end{aligned}
\]

Men tend to have the higher course evaluation for the same beauty rating.

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.

```{r}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

R creates dummy variables numbering (category levels - 1). The unnamed level is a reference variable.

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

**ANSWER**

I expect the `cls` variables not to have an association with the professor score. Therefore: `cls_perc_eval`, `cls_students`, `cls_level`, `cls_profs`, or `cls_credits`. To pick one, I guess `cls_credits`. But I always thought there was a problem with the design including some classes with multiple instructors, so `cls_profs` would be a close second for my guess.

Let's run the model...

```{r m_full, eval = TRUE, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.
    
**ANSWER**

Output: cls_profssingle has a P value of 0.77806.

So `cls_profs` is it.

13. Interpret the coefficient associated with the ethnicity variable.

**ANSWER**

All other predictors being equal, an additional point in beauty rating increases the score of someone who is not a minority by 0.12.

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?

```{r}
m_full_less_1 <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full_less_1)
```

There were some very slight shifts, but no major upset, revealing collinearity with the other predictor variables.

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.

**ANSWER**

Full model adjusted $R^2$: 0.1617.

`cls_profsingle` has highest P at 0.77806.

Exclude `cls_prof`, adjusted $R^2$: 0.1634.

`cls_levelupper` has highest P at 0.292200.

Exclude `cls_level`, adjusted $R^2$: 0.1632.

That reduction reduced the predictive value of the model, so we keep `cls_level`. This is the same model from a couple questions back.

```{r}
m_full_less_1 <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full_less_1)
```

$\widehat{score}$ = 4.0872523 - 0.1476746 * $rank_{tenure track}$ - 0.0973829 * $rank_{tenured}$ + 0.1274458 * $ethnicity_{not minority}$ + 0.2101231 * $gender_{male}$ - 0.2282894 * $language_{non-english}$ - 0.0089992 * $age$ + 0.0052888 * $cls\_perc\_eval$ + 0.0004687 * $cls\_students$ + 0.0606374 * $cls\_level_{upper}$ + 0.5061196 * $cls\_credits_{one credit}$ + 0.0398629 * $bty\_avg$- 0.1083227 * $pic\_outfit_{not formal}$ - 0.2190527 * $pic\_color_{color}$

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.

**ANSWER**

Seems reasonable. No outliers. Linear. Consistent variability.

**Check for outliers**
```{r}
hist(m_full_less_1$residuals)
```

**Linearity and variability**  
```{r}
plot(m_full_less_1$residuals ~ evals$bty_avg)
abline(h = 0, lty = 3)
```

17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?
    
**ANSWER**

If it is a sample of professors instead of courses, then I think the scores should have been weighted by the number of courses each professor taught. Varibility in the number of courses taught would bias the results. 

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.
    
**ANSWER**

Instructor features.

Rank: Teaching (not tenured or tenure track).

Ethnicity: Not a minority.

Gender: Male.

Language: English.

Age: Younger the better.

Beauty: The more attractive the better.

Photo: Color and formal attire.

Class features.

More evaluations the better.

Higher class size the better.

Upper division.

One-credit class.

19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?

**ANSWER**

No. mostly due to differences in studentbody demographics. A school somewhere other than Texas may have a higher proportion of minority students and the coefficient for non-minorities may be higher.