---
title: "Web APIs"
author: "Jai Jeffryes"
date: "10/21/2019"
output:
  html_document:
    code_download: yes
    code_folding: hide
    highlight: pygments
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(jsonlite)
library(stringr)
library(tibble)
library(tidyr)
library(mongolite)

nytimes_api_key <- readLines("nytimes_key.txt")
```

## URL
Some JSON calls require a query string, others don't. Use a function to build the URL so you don't drive yourself crazy with positioning of `?` and `&`.

```{r}
get_nyt_json <- function(api_key, api_url, query = NULL) {
	if (is.null(query)) {
		base_url <- str_c(api_url, "?api-key=", nytimes_api_key)
	} else {
		base_url <- str_c(api_url, "?", query, "&api-key=", nytimes_api_key)
	}
	
	print(base_url)
	json <- fromJSON(base_url)
	json
}
```

## Go get 'em
I tried three retrieval scenarios.

1. Simple list. Books API.
1. Nested list of data frames. Books API.
1. Query. Movie reviews API.

The `flatten` parameter in `jsonlite::fromJSON()` and the function`jsonlite::flatten()` function achieved nothing with these JSON documents. However, `unnest()` from the `tidyr` package did flatten the nested data frames I wanted to see in the book overview.

### Don't do me so many favors, `tibble()`
However, if you want to flatten with `tidyr::unnest()`, use base data frames, not a `tibble`. A base data frame gathers the nested lists into columns. Ironically, `as.tibble()` does not.

```{r}
# Get best sellers list names.
api_url <- "https://api.nytimes.com/svc/books/v3/lists/names.json"
book_list <- get_nyt_json(nytimes_api_key, api_url)
book_list_df <- as_tibble(book_list)
head(book_list_df)

# Get best sellers overviews.
# Top 5 books for all lists
api_url <- "https://api.nytimes.com/svc/books/v3/lists/overview.json"
book_overview <- get_nyt_json(nytimes_api_key, api_url)
book_overview_df <- data.frame(book_overview) %>% unnest(results.lists.books)
head(book_overview_df)
# Call out the flattened titles.
head(book_overview_df[, c("title", "contributor", "publisher")])

# Movie reviews
api_url <- "https://api.nytimes.com/svc/movies/v2/reviews/search.json"
query <- "query=hollywood"
movie_review <- get_nyt_json(nytimes_api_key, api_url, query)
movie_review_df <- as_tibble(movie_review)
head(movie_review_df)
```

## MongoDB
I wanted to take advantage of this chance to get started with MongoDB.

```{r}
# Create files of collections
cat(str_c("Creating files...\n"))

stream_out(x = book_list_df, file("book_list.json"))
stream_out(x = book_overview_df, file("book_overview.json"))
stream_out(x = movie_review_df, file("movie_review.json"))

# Import a collection and report action.
m_coll <- function(collection, db, fil) {
	m_con <- mongo(collection = collection, db = db)
	
	cat("Dropping collection if it is there...\n")
	m_con$drop()
	print(m_con$count())
	
	cat(str_c("Importing collection: ", collection, "\nDatabase: ", db, "\nFile: ", fil, "\n"))
	m_con$import(file(fil))
	cnt <- m_con$count()
	cat(str_c("Records inserted: ", cnt, "\n"))
		
	m_con$disconnect()
}

m_coll("book_list", "nytimes", "book_list.json")
#m_coll("book_overview", "nytimes", "book_overview.json") # Clean the keys.
m_coll("movie_review", "nytimes", "movie_review.json")
```

