---
title: "Chapter 5, Page 200"
author: "Jai Jeffryes"
date: "3/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 5.1. Problem 22.
Assume that we are making raisin cookies. We put a box of 600 raisins into our dough mix, mix up the dough, then make from the dough 500 cookies. We then ask for the probability that a randomly chosen cookie will have 0, 1, 2, ... raisins. Consider the cookies as trials in an experiment, and let $X$ be the random variable which gives the number of raisins in a given cookie. Then we can regard the number of raisins in a cookie as the result of $n = 600$ independent trials with probability $p = 1/500$ for success on each trial. Since $n$ is large and $p$ is small, we can use the Poisson approximation with $\lambda = 600(1/500) = 1.2$. Determine the probability that a given cookie will have at least five raisins.

### Answer
I'm adopting the approach outlined in Example 5.4 on page 191. Problem 22 deals with raison cookies instead of bombs. We approximate the probability of getting $j$ raisons using the Poisson distribution.

$\lambda = 1.2$

$p(j) = e^{-\lambda}\lambda^j/j!$

I think I want to calculate the probability for 1, 2, 3, and 4 raisons. Then subtract that sum from 1.
```{r}
p_1to4 <- 0
lambda = 1.2

for(j in 1:4) {
	p_1to4 = p_1to4 + (exp(-lambda) * lambda^j) / factorial(j)
}

1 - p_1to4
```

However, that doesn't look at all right to me, so I'm turning in what I have and it's back to the drawing board in the morning.