---
title: "DATA605 Discussion Week 4"
subtitle: "Page 477, Exercise C20"
author: "Jai Jeffryes"
date: "2/16/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## C20
Example **SAR** concludes with an expression for a vector $u \in \mathbb{C}^5$ that we believe will create the vector $v \in \mathbb{C}^5$ when used to evaluate $T$. That is, $T(u) = v$. Verify this assertion by actually evaluating $T$ with $u$. If you do not have the patience to push around all these symbols, try choosing a numerical instance of $v$, compute $u$, and then compute $T(u)$, which should result in $v$.

### Given
```{r}
# Archetype R
R <- matrix(c(
	-65, 128, 10, -262, 40,
	36, -73, -1, 151, -16,
	-44, 88, 5, -180, 24,
	34, -68, -3, 140, -18,
	12, -24, -1, 49, -5
), nrow = 5, byrow = T)
R

# Beezer assertion. R is non-singular and has an inverse.
R.inv <- solve(R)
R.inv
```

### Verify
- Choose numerical instance of $v$
```{r}
v <- c(22, -3, 100, 5, 17)
v
```

- Compute $u$
```{r}
u <- R.inv %*% v
u
```

- Compute $T(u)$ and compare
```{r}
t.u <- R %*% u
t.u
```

- Compare the two results
```{r}
# Here is v.
v

# Here is T(u). Expected: same as v.
t.u
```

## Conclusion
Expected result is confirmed. Archetype **R** is a surjective linear transformation.
